---
title: "HUVEC_ZEB1_fullWorkflow"
author: "Joseph L. Horder"
date: '2022-07-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the packages that are required for the analysis.

```{r}
library(DESeq2)
library(biomaRt)
library(PCAtools)
library(ggvenn)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ComplexHeatmap)
library(stringr)
library(dplyr)
library(tidytext)
library(tidyverse)
library(ggpubr)
library(msigdbr)
```

Prepare the input files. A meta file needs to be created that contains any information that is relevant to the analysis.
This includes sample name, experimental group, cell type, treatment, sample ID, etc.
The required input files are metadata and quantified readcounts.

```{r}
# A variable can be set for the file path to save having to continually write it out.
file_path <- '~/cnv_paper/'
```

```{r}
# Specify the meta data file, which is named 'groups.csv' here.
sample_info <- file.path(file_path, 'data/bulkrnaseq/groups.csv')
# sample_info <- file.path(file_path, 'data/bulkrnaseq/whole_test/groups.csv')

# Read the groups.csv file and store as a variable named meta.
# The rownames are set to the first column of the data.
# stringsAsFactors specifies whether strings (text) are treated as factor variables or just strings.
meta <- read.csv(sample_info, row.names = 1, stringsAsFactors = FALSE)
```

```{r}
# Specify the readcounts data file, which is named 'counts_processed.csv' here.
readcounts <- file.path(file_path, 'data/bulkrnaseq/counts_processed.csv')
# readcounts <- file.path(file_path, 'data/bulkrnaseq/whole_test/counts_processed.csv')

# Read the readcount.csv file and store as a variable named data.
# The header argument specifies wether to top row of the data is a header (column names).
data <- read.csv(readcounts, row.names = 1, stringsAsFactors = FALSE, header = TRUE)
```

It is essential that the order of the rownames in the meta data and the column names in the readcounts data match.
Otherwise, readcounts may be assigned to the wrong sample and cause false results.

```{r}
# Ensure that the meta data ONLY includes rows (sample info) for the samples (columns) that are in the readcount data file.
meta <- meta[rownames(meta) %in% colnames(data),]

# Reorder the columns of the readcounts data to match the row order of the meta data.
data <- data[,rownames(meta)]
```

DESeq2.

The first step is to build a DESeq2 object. This is the framework upon which the analysis will be built.

```{r}
# Create the DESeq2 object specifying the readcount data as countData, the meta data as colData and the design for the DESeq2 object as the group column of the meta data.
# Different kinds of objects can be built for different analysis situations. See the README.md.
# As we are using data matrices for the object with use the DESeqDataSetFromMatrix() function.
dds <- DESeqDataSetFromMatrix(countData = data,
                              colData = meta,
                              design = ~ group)
```

Generating the normalised counts by DESeq2's median of ratios method.

```{r}
# First the size factors need to be calculated. These are the factors to accounts for library depth and RNA composition.
dds <- estimateSizeFactors(dds)

# Print the size factors for visual inspection.
sizeFactors(dds)
```

Retrieving the normalised counts from the DESeq2 object to save as a separate variable.

```{r}
# Access normalised counts.
normalised_counts <- counts(dds, normalized = TRUE)

# Coerce the normalised counts into a dataframe.
normalised_counts <- as.data.frame(normalised_counts)
```

BioMart. 
The normalised readcounts have the genes annotated in the ensembl format (ENSG).
Whilst it is important to keep using the ensembl names throughout the analysis (avoids ambiguity, potential duplicates, loss of information etc), they are not necessarily the most informative.
We can match the ensembl names to gene symbols using BioMart.

```{r}
# First, create a 'mart object'.
mart_data <- useEnsembl("genes", dataset = "hsapiens_gene_ensembl")

# Retrieve gene symbol information for the ensembl gene ids in normalised counts.
ens_to_symbol <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
                       values = rownames(data),
                       filters = "ensembl_gene_id",
                       mart = mart_data)
```

Adding the gene symbol information to the normalised counts.

```{r}
normalised_counts$symbol <- ens_to_symbol$hgnc_symbol[match(rownames(normalised_counts), ens_to_symbol$ensembl_gene_id)]

normalised_counts$gene_biotype <- ens_to_symbol$gene_biotype[match(rownames(normalised_counts), ens_to_symbol$ensembl_gene_id)]
```

Extracting the normalised counts to their own .csv file.
The normalised counts are those that will be used for all downstream analyses.
It is useful to have these in a .csv file that can be easily viewed, shared and used in other programs/webservers.

```{r}
write.csv(as.data.frame(normalised_counts), 
           file=file.path(file_path,"output/data/19_HUVEC_normalised_counts.csv"))
```

Regularised log transformation.
This transforms the data to allow for sample QA, PCA, heatmaps etc to prevent skewing.
The blind argument means the transformation is fully unsupervised and blind to sample.

```{r}
rld <- rlog(dds, blind = TRUE)
```

PCA analysis.
PCA analysis is used to assess variation in the data.

```{r}
# Coerce the rlog data into a matrix.
rld_mat <- assay(rld)
```

```{r}
# Run the PCA.
pca_output <- pca(rld_mat, metadata = meta)
```

Plot the PCA for all samples and cell types in a biplot.

```{r}
# Generate the plot.
pca1_2_huvec <- biplot(pca_output,
                 colby = 'treatment',
                 legendPosition = 'right',
                 #colLegendTitle = 'Cell Type',
                 shapeLegendTitle = 'Treatment',
                 lab = meta$ID)
# Show the plot.
pca1_2_huvec

ggsave(file=file.path(file_path,"output/plots/18_pca_huvec.svg"), height = 5, width = 6)
```

Plotting a scree plot. This allows us to see the total number of principal components and the variance of each principal component.

```{r}
# Scree plot for all samples.
scree <- screeplot(pca_output)

# Show the scree plot.
scree
```

Differential expression analysis.

Set the factor level to the appropriate experimental group for log2FC shrinking.

```{r}
dds$group <- relevel(dds$group, ref = "HUVEC_control")
```

Now the DE analysis can be run with a simple function.

```{r}
dds <- DESeq(dds)
```

State the contrast of data to extract results from.

```{r}
# State the contrast.
contrast <- c("group", "HUVEC_ZEB1_KD", "HUVEC_control")
```

Extract the DE results with shrunk log2FC.

```{r}
# Gather the results from the specified contrast.
# alpha refers to the BH adjusted p-value used for significance. The results will try to fit to this p-value.
results <- results(dds, contrast = contrast, alpha = 0.05)

# Check the names of the coefficients that can be used in the log2FC shrinking.
resultsNames(dds)

# Extract the shrunk results. apeglm is the type of shrinkage. See DESeq2 documentation for more detail.
results <- lfcShrink(dds,
                     res = results,
                     coef = "group_HUVEC_ZEB1_KD_vs_HUVEC_control", 
                     type="apeglm")

# Print a summary of the results.
summary(results)
```

Add gene symbols to the DE results.

```{r}
results$symbol <- ens_to_symbol$hgnc_symbol[match(rownames(results), ens_to_symbol$ensembl_gene_id)]
```

Extracting significant DEGs based on thresholds.

Set threshold values.

```{r}
# Set the fold change threshold.
fc_thresh = log2(1.5)

# Set the adjusted p-value threshold.
padj_thresh = 0.05
```

Retrieve all DE results.

```{r}
# Retrieve all DE results, regardless of thresholds. Ordered by padj.
all_DE <- results[order(results$padj),]

# Export to a .csv
write.csv(as.data.frame(all_DE),
          file = file.path(file_path, "output/data/20_all_DE.csv"))
```

Retrieve all significant DE results (padj < 0.05).

```{r}
# All DE results with padj < 0.05.
padj_DE <- subset(results, padj < padj_thresh)

# Order results by padj.
padj_DE <- padj_DE[order(padj_DE$padj),]

# Export to a .csv
write.csv(as.data.frame(padj_DE),
          file = file.path(file_path, "output/data/21_padj_DE.csv"))
```

Retrieve all significant DE results (padj < 0.05) and log2FC threshold 1.5.

```{r}
# All DE results with padj < 0.05.
fc_DE <- subset(results, padj < padj_thresh & abs(log2FoldChange) >= fc_thresh)

# Order results by log2FoldChange.
fc_DE <- fc_DE[order(-fc_DE$log2FoldChange),]

# Export to a .csv
write.csv(as.data.frame(fc_DE),
          file = file.path(file_path, "output/data/22_log2fc_padj_DE.csv"))
```

```{r}
normCounts <- normalised_counts[1:6]
normCounts

normCounts_noZero <- normCounts[!(rowSums(normCounts) == 0),]
normCounts_noZero

reference <- rownames(normCounts_noZero)
reference

write.csv(as.data.frame(reference), 
            file=file.path(file_path,"output/data/23_universe_reference.csv"))
```

Plot Zeb1 normalised counts bar chart

```{r}
zeb1_counts <- as.data.frame(t(normCounts[rownames(normCounts) == "ENSG00000148516",]))
zeb1_counts$group <- c("NS_control", "NS_control", "NS_control", "ZEB1_KD", "ZEB1_KD", "ZEB1_KD")

# (Optional) Compute summary statistics for reference
summary_data <- zeb1_counts %>%
  group_by(group) %>%
  summarise(mean_count = mean(ENSG00000148516),
            sd_count = sd(ENSG00000148516))
print(summary_data)

# significance with t-test
t.test(ENSG00000148516 ~ group, data = zeb1_counts, var.equal = FALSE)

# stat_summary with geom="bar" plots the mean for each group.
# stat_summary with fun.data=mean_sdl (mult=1) adds error bars for ±1 SD.
# geom_jitter overlays individual data points.
ggplot(zeb1_counts, aes(x = group, y = ENSG00000148516)) +
  stat_summary(fun = mean, geom = "bar", fill = "darkgrey", width = 0.3) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "errorbar", width = 0.2) +
  stat_compare_means(
    method = "t.test",
    label = "p.format",
    comparisons = list(c("NS_control", "ZEB1_KD")),
    var.equal = FALSE
  ) +
  geom_jitter(width = 0.1, size = 1, color = "black") +
  labs(x = "Group", y = "DESeq2 normalized count", title = "ZEB1 Expression") +
  theme_classic()
ggsave(file = file.path(file_path, "output/plots/19_Zeb1_norm_counts.svg"), width = 3, height = 4)
```

Volcano for all significant

```{r}
# Specify the threshold.
fc_thresh <- 0

# The results need to be in a dataframe format.
results_15 <- as.data.frame(results)

# Specify the category of DEGs depending on the threshold.
results_15$direction <- "Not Differentially Expressed"
results_15$direction[results_15$log2FoldChange >= fc_thresh & results_15$padj < padj_thresh] <- "Upregulated"
results_15$direction[results_15$log2FoldChange <= -fc_thresh & results_15$padj < padj_thresh] <- "Downregulated"

# Label data for ZEB1 only
ZEB1_labels <- "ZEB1"
ZEB1_labels_data <- results_15[match(ZEB1_labels, results_15$symbol), ]


# Volcano plot
ggplot(data = results_15, aes(log2FoldChange, -log10(padj))) +
  geom_point(aes(col = direction), size = 0.5) +
  
  # Colour scale for DEGs
  scale_colour_manual(
    name = "Differential Expression",
    values = c("Upregulated" = "red", 
               "Downregulated" = "blue", 
               "Not Differentially Expressed" = "grey")
  ) +
  
  # Threshold lines
  geom_vline(xintercept = c(fc_thresh, -fc_thresh), linetype = "dashed") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed") +
  
  # Label only ZEB1
  geom_text_repel(
    data = ZEB1_labels_data,
    aes(log2FoldChange, -log10(padj), label = symbol),
    fontface = "bold.italic",
    size = 2,
    min.segment.length = unit(0, "lines"),
    force = 10,
    max.iter = Inf
  ) +
  
  ggtitle("HUVEC ZEB1 KD Volcano Plot") +
  theme_classic()

ggsave(file = file.path(file_path, "output/plots/20_volcano_all_padj_labels.svg"), width = 5, height = 3)
```

Gene Ontology analysis.
Use the custom background reference!

```{r}
HUVECego_padj <- enrichGO(
    gene          = rownames(padj_DE),
    OrgDb         = org.Hs.eg.db,
    keyType       = "ENSEMBL",
    ont           = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    readable      = TRUE,
    universe = reference
)

simpleHUVECZEBego_padj <- clusterProfiler::simplify(HUVECego_padj, cutoff = 0.7, by = "p.adjust", select_fun = min)

write.csv(simpleHUVECZEBego_padj@result, file = file.path(file_path, "output/data/24_HUVEC_zeb1_DE_GO_padj.csv"))
```

```{r}
zeb1_DE_genes_entrez <- bitr(rownames(padj_DE), fromType = "ENSEMBL", toType = "ENTREZID",
                             OrgDb = "org.Hs.eg.db")

reference_entrez <- bitr(reference, fromType = "ENSEMBL", toType = "ENTREZID",
                             OrgDb = "org.Hs.eg.db")

ekeggZeb1DE <- enrichKEGG(
    gene = zeb1_DE_genes_entrez$ENTREZID,
    organism      = 'hsa',
    keyType       = "kegg",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    universe = reference_entrez$ENTREZID)

ekeggZeb1DE <- setReadable(ekeggZeb1DE, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

ekeggZeb1DE <- ekeggZeb1DE@result %>%
  dplyr::filter(qvalue < 0.05) %>%
  arrange(qvalue)
head(ekeggZeb1DE)

write.csv(ekeggZeb1DE, file = file.path(file_path, "/output/data/25_HUVEC_zeb1_DE_KEGG.csv"))
```

```{r}
# Perform functional enrichment analysis of DR genes against HALLMARKS
# Retrieve the sets of hallmarks data. Refine the data.
m_t2g <- msigdbr(species = "Homo sapiens", collection = "H") %>% 
  dplyr::select(gs_name, ensembl_gene)
head(m_t2g)

deg_hallmarks <- enricher(rownames(padj_DE), 
                         TERM2GENE = m_t2g, 
                         qvalueCutoff = 0.05, 
                         universe = reference)
deg_hallmarks <- deg_hallmarks@result %>%
  dplyr::filter(qvalue < 0.05) %>%
  arrange(qvalue)
head(deg_hallmarks)

# Create a lookup vector from padj_df
id_to_symbol <- padj_df$symbol
names(id_to_symbol) <- rownames(padj_df)

# Define a function to convert Ensembl IDs in geneID to symbols
convert_geneIDs <- function(geneID_str) {
  ids <- unlist(strsplit(geneID_str, "/"))        # split Ensembl IDs by "/"
  symbols <- id_to_symbol[ids]                    # map to gene symbols
  symbols <- symbols[!is.na(symbols)]             # remove missing ones
  paste(symbols, collapse = "/")                  # rejoin into a string
}

# Apply conversion to dr_hallmarks
deg_hallmarks$symbol <- sapply(deg_hallmarks$geneID, convert_geneIDs)

write.csv(deg_hallmarks, file = "../output/data/26_HUVEC_zeb1_DE_HALLMARKS.csv")
```

```{r}
# Dotplot of combined functional enrichment results for inflammation-related results
# First, combine the data together.
GO <- as.data.frame(simpleHUVECZEBego_padj@result)
KEGG <- as.data.frame(ekeggZeb1DE)
HALL <- as.data.frame(deg_hallmarks)

GO$Type <- "GO"
KEGG$Type <- "KEGG"
HALL$Type <- "HALLMARKS"

# Concatenate ID and Description for GO and KEGG
GO <- GO %>%
  mutate(Description = paste0(Description, " ", "(", ID, ")"))

KEGG <- KEGG %>%
  mutate(Description = paste0(Description, " ", "(", ID, ")"))
KEGG <- KEGG[, !(names(KEGG) %in% c("category", "subcategory"))]

all_res <- bind_rows(GO, KEGG, HALL)

selected_terms <- c("HALLMARK_INTERFERON_GAMMA_RESPONSE",
"HALLMARK_INTERFERON_ALPHA_RESPONSE",
"HALLMARK_INFLAMMATORY_RESPONSE",
"HALLMARK_TNFA_SIGNALING_VIA_NFKB",
"TNF signaling pathway (hsa04668)",
"Cytokine-cytokine receptor interaction (hsa04060)",
"IL-17 signaling pathway (hsa04657)",
"NF-kappa B signaling pathway (hsa04064)",
"Chemokine signaling pathway (hsa04062)",
"Leukocyte transendothelial migration (hsa04670)",
"AGE-RAGE signaling pathway in diabetic complications (hsa04933)",
"NOD-like receptor signaling pathway (hsa04621)",
"cell chemotaxis (GO:0060326)",
"cellular response to tumor necrosis factor (GO:0071356)",
"chemokine production (GO:0032602)",
"chemotaxis (GO:0006935)",
"cytokine-mediated signaling pathway (GO:0019221)",
"interferon-mediated signaling pathway (GO:0140888)",
"leukocyte cell-cell adhesion (GO:0007159)",
"leukocyte migration (GO:0050900)",
"leukocyte proliferation (GO:0070661)",
"negative regulation of interleukin-1 production (GO:0032692)",
"neutrophil activation (GO:0042119)",
"regulation of chemokine production (GO:0032642)",
"regulation of chemotaxis (GO:0050920)",
"regulation of leukocyte migration (GO:0002685)",
"regulation of leukocyte proliferation (GO:0070663)",
"regulation of response to cytokine stimulus (GO:0060759)",
"regulation of response to wounding (GO:1903034)",
"response to chemokine (GO:1990868)",
"response to interferon-alpha (GO:0035455)",
"response to interferon-beta (GO:0035456)",
"response to tumor necrosis factor (GO:0034612)",
"response to type II interferon (GO:0034341)",
"T cell activation (GO:0042110)",
"T cell activation involved in immune response (GO:0002286)",
"wound healing (GO:0042060)")

# Filter selected pathways by Description
plot_data <- all_res %>%
  filter(Description %in% selected_terms)

# Create readable labels
plot_data <- plot_data %>%
  mutate(Description = factor(Description, levels = rev(unique(Description))))

# Convert GeneRatio to numeric
plot_data <- plot_data %>%
  separate(GeneRatio, into = c("x", "y"), sep = "/") %>%
  mutate(GeneRatioNumeric = as.numeric(x) / as.numeric(y))

# Ensure key columns are numeric
plot_data <- plot_data %>%
  mutate(
    qvalue = as.numeric(qvalue),
    Count    = as.numeric(Count))

# Order the y axis by significance for each database
plot_data <- plot_data %>%
  arrange(qvalue, Description) %>%
  mutate(Description = factor(Description, levels = rev(unique(Description))))

# Make dotplot
p <- ggplot(
  plot_data,
  aes(x = GeneRatioNumeric, y = Description, size = Count, color = qvalue)
) +
  geom_point() +
  scale_color_gradient(high = "#327eba", low = "#e06663", name = "qvalue") +
  guides(color = guide_colorbar(reverse = TRUE)) +  # red = more significant at top of legend
  scale_size_continuous(breaks = c(5,10,15,20,25,30), range=c(2,8)) +
  labs(x = "Gene Ratio", y = NULL, title = "") +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10)
  )

print(p)
ggsave(
  plot = p,
  filename = "../output/plots/21_HUVEC_zeb1_KD_funenrich_inflamm.svg",
  width = 20,
  height = 20,
  units = "cm",
  dpi = 320
)
```
